---
title: "Stat380_MiniProject1"
author: " "
format: html
editor: visual
---

## Importing and Joining Data

```{r}
library(readr)
library(tidyverse)
#Predicted Variable
pop <- read_csv("data/pop.csv")

#Predictor Variables
sanitation <- read_csv("data/at_least_basic_sanitation_overall_access_percent.csv")
fertility <- read_csv("data/children_per_woman_total_fertility.csv")
co2 <- read_csv("data/co2_pcap_cons.csv")
happiness <- read_csv("data/hapiscore_whr.csv")
income <- read_csv("data/ilevels4_wb.csv")
poverty <- read_csv("data/number_of_people_in_poverty.csv")
population <- read_csv("data/pop.csv")

# Joining Tables
drop_and_pivot <- function(df, value_name) {
  df |>
    select(-any_of(as.character(2023:2100))) |>
    pivot_longer(
      cols = -c(geo, name),
      names_to = "year",
      values_to = value_name
    ) |>
    mutate(year = as.integer(year))
}

population_long <- drop_and_pivot(population, "pop")
sanitation_long <- drop_and_pivot(sanitation, "sanitation_access_pct")
fertility_long  <- drop_and_pivot(fertility, "children_per_woman")
co2_long        <- drop_and_pivot(co2, "co2_per_capita")
happiness_long  <- drop_and_pivot(happiness, "happiness_score")
income_long     <- drop_and_pivot(income, "income_level")
poverty_long    <- drop_and_pivot(poverty, "people_in_poverty")
```

## Join into a single dataset

```{r}
#| label: join
combined <- list(
  sanitation_long, fertility_long, co2_long,
  happiness_long, income_long, poverty_long, population_long
) |>
  reduce(full_join, by = c("geo", "name", "year")) |>
  arrange(geo, year)

glimpse(combined)
```

```{r}
#| label: check
range(combined$year)
n_distinct(combined$geo)
head(combined)
```

```{r}
# Summary Statistics and Visualization
# Number of People In Poverty (Quantitative)
poverty_long |>
  group_by(name) |>
  summarize(mean_poverty = mean(people_in_poverty, na.rm = TRUE),
    min_poverty = min(people_in_poverty, na.rm = TRUE),
    max_poverty = max(people_in_poverty, na.rm = TRUE),
    median_poverty = median(people_in_poverty, na.rm = TRUE),
    sd_poverty = sd(people_in_poverty, na.rm = TRUE))
poverty_long |> 
  ggplot(aes(x=year, y=people_in_poverty, color = name)) +
  geom_line(linewidth = .001) +
  labs(
    title = "Poverty Counts Over Time",
    x = "Year",
    y= "Number of People in Poverty"
  ) +
  theme(legend.text = element_text(size = 8))

#Basic Sanitation Overall (Quantitative)
sanitation_long |>
  group_by(name) |>
  summarize(mean_sanitation_percentage = mean(sanitation_access_pct, na.rm = TRUE),
    min_sanitation_percentage = min(sanitation_access_pct, na.rm = TRUE),
    max_sanitation_percentage = max(sanitation_access_pct, na.rm = TRUE),
    median_sanitation_percentage = median(sanitation_access_pct, na.rm = TRUE),
    sd_sanitation_percentage = sd(sanitation_access_pct, na.rm = TRUE))

sanitation_long |>
  ggplot(aes(x=year, y=sanitation_access_pct)) +
  geom_line(linewidth = .001) +
  labs(
    title = "Basic Sanitation Percentage Over Time",
    x = "Year",
    y = "Sanitation Access Percentage"
  ) +
  facet_wrap(vars(name))
```
Observations of Poverty: There were many NA values inside of poverty specifically for countries that were made quite recently. These will have to be removed as they will not work when creating a model.

Observations of Sanitation: The data is from 2000 beyond and contains some NA values which will have to be removed later. A notable country in this data set is Andorra as it has 100% sanitation access for all the years recorded. 

```{r}
set.seed(123)
sample <- sample(c(TRUE,FALSE), nrow(combined), 
                 replace=TRUE, prob=c(0.8,0.2))

# creating training dataset
train_dataset  <- combined[sample, ]

# creating testing dataset
validation_dataset  <- combined[!sample, ]

Model3 = train_dataset |>
  lm(pop~sanitation_access_pct+people_in_poverty, data= _) #base R code, the _ tells R where to pipe
summary(Model3)
```
