---
title: "CO2 Emissions & Income Groups Predictors"
format:
  pdf:
    fig-height: 3
    fig-width: 6
editor: visual
---

> **Source disclosure**
>
> Reference sources for the techniques used:
>
> - Visualization: ggplot2 reference for `geom_jitter`, <https://ggplot2.tidyverse.org/reference/geom_jitter.html>
> - Cleaning: R for Data Science, data tidying chapter, <https://r4ds.hadley.nz/data-tidy.html>
> - Other: R Markdown Cookbook, tables with `kable`, <https://bookdown.org/yihui/rmarkdown-cookbook/kable.html>

## Libraries

```{r}
#| message: false
library(tidyverse)
```

## Import the data files

Each file is in wide format with one row per country and one column per year.

```{r}
#| message: false
sanitation <- read_csv("../data/predictors/at_least_basic_sanitation_overall_access_percent.csv")
fertility  <- read_csv("../data/predictors/children_per_woman_total_fertility.csv")
co2        <- read_csv("../data/predictors/co2_pcap_cons.csv")
happiness  <- read_csv("../data/predictors/hapiscore_whr.csv")
income     <- read_csv("../data/predictors/ilevels4_wb.csv", col_types = cols(.default = "c"))
poverty    <- read_csv("../data/predictors/number_of_people_in_poverty.csv")
```

## Dropping Year Columns

The year columns from 2023 onwards are dropped first. Each dataset is then changed to long format so that all of them share the same geo, name and year keys, which avoids duplicated year column names when joining.

```{r}
drop_and_pivot <- function(df, value_name) {
  df |>
    select(-any_of(as.character(2023:2100))) |>
    pivot_longer(
      cols = -c(geo, name),
      names_to = "year",
      values_to = value_name
    ) |>
    mutate(year = as.integer(year))
}

sanitation_long <- drop_and_pivot(sanitation, "sanitation_access_pct")
fertility_long  <- drop_and_pivot(fertility, "children_per_woman")
co2_long        <- drop_and_pivot(co2, "co2_per_capita")
happiness_long  <- drop_and_pivot(happiness, "happiness_score")
income_long     <- drop_and_pivot(income, "income_level")
poverty_long    <- drop_and_pivot(poverty, "people_in_poverty")
```

## Join into a single dataset

```{r}
combined <- list(
  sanitation_long, fertility_long, co2_long,
  happiness_long, income_long, poverty_long
) |>
  reduce(full_join, by = c("geo", "name", "year")) |>
  arrange(geo, year)

glimpse(combined)
```

```{r}
range(combined$year)
n_distinct(combined$geo)
head(combined)
```

# Ian Peralta: CO2 Emissions and Income Groups

Population, the response, is not yet in `combined`, so it is imported here.

```{r}
#| message: false
population      <- read_csv("../data/response/pop.csv")
population_long <- drop_and_pivot(population, "population")
```

## CO2 Emissions per Capita: Consumption-Based

`co2_per_capita` is quantitative and continuous, measured in tonnes of CO2 per person per year. Values run from near zero to several tens of tonnes.

```{r}
co2_long |>
  summarise(
    n           = n(),
    n_missing   = sum(is.na(co2_per_capita)),
    pct_missing = round(100 * mean(is.na(co2_per_capita)), 1),
    first_year  = min(year[!is.na(co2_per_capita)]),
    last_year   = max(year[!is.na(co2_per_capita)])
  ) |>
  knitr::kable()
```

About 4% of records are missing. Most of this is structural, since many countries are only reported from later years.

```{r}
co2_long |>
  filter(co2_per_capita < 0) |>
  arrange(co2_per_capita) |>
  knitr::kable(digits = 3)
```

Seven values from Panama are slightly negative. Consumption-based accounting subtracts exported emissions, so a trade hub can go net negative. These are kept as a real feature of the method, but they matter for modeling.

```{r}
co2_latest <- co2_long |>
  filter(!is.na(co2_per_capita), year == max(year[!is.na(co2_per_capita)]))

co2_long |>
  filter(!is.na(co2_per_capita)) |>
  summarise(mean = mean(co2_per_capita), median = median(co2_per_capita),
            sd = sd(co2_per_capita), min = min(co2_per_capita),
            max = max(co2_per_capita)) |>
  knitr::kable(digits = 2, caption = "CO2 per capita, all years")

co2_latest |>
  summarise(year = unique(year), n_countries = n(), mean = mean(co2_per_capita),
            median = median(co2_per_capita), sd = sd(co2_per_capita),
            min = min(co2_per_capita), max = max(co2_per_capita)) |>
  knitr::kable(digits = 2, caption = "CO2 per capita, 2022")
```

```{r}
#| fig-cap: "Mean CO2 emissions per capita by year"
co2_long |>
  filter(!is.na(co2_per_capita)) |>
  group_by(year) |>
  summarise(mean_co2 = mean(co2_per_capita)) |>
  ggplot(aes(year, mean_co2)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  labs(x = "Year", y = "Mean CO2 per capita, tonnes") +
  theme_minimal()
```

```{r}
#| fig-cap: "Distribution of CO2 emissions per capita, 2022"
ggplot(co2_latest, aes(co2_per_capita)) +
  geom_histogram(bins = 25, fill = "blue", color = "white") +
  labs(x = "CO2 per capita, tonnes", y = "Number of countries") +
  theme_minimal()
```

```{r}
bind_rows(
  co2_latest |> slice_max(co2_per_capita, n = 5) |> mutate(group = "Highest"),
  co2_latest |> slice_min(co2_per_capita, n = 5) |> mutate(group = "Lowest")
) |>
  select(group, name, co2_per_capita) |>
  knitr::kable(digits = 2, caption = "Highest and lowest CO2 per capita, 2022")
```

The 2022 distribution is right-skewed, with most countries below 5 tonnes. The highest are Qatar, Singapore, Brunei, Kuwait and the UAE, small wealthy economies tied to oil, gas or trade. The lowest are the DR Congo, Burundi, Somalia, the Central African Republic and Chad. Average emissions climbed through industrialization and have leveled off since about 1980.

```{r}
#| fig-cap: "CO2 emissions per capita against population, 2022"
co2_pop_2022 <- co2_latest |>
  inner_join(population_long |> filter(year == 2022),
             by = c("geo", "name", "year"))

cor(co2_pop_2022$co2_per_capita, log10(co2_pop_2022$population))

ggplot(co2_pop_2022, aes(population, co2_per_capita)) +
  geom_point(alpha = 0.6, color = "blue") +
  scale_x_log10(labels = scales::comma) +
  labs(x = "Population, log scale", y = "CO2 per capita, tonnes") +
  theme_minimal()
```

There is almost no relationship between population and per-capita emissions, with r = -0.06 against log population. This is expected, since the measure is already per person and depends more on income and energy use than country size.

## 4 Income Groups: World Bank

`income_level` is categorical and ordinal with four levels: Low, Lower middle, Upper middle and High income, assigned by the World Bank from GNI per capita.

```{r}
income_long <- income_long |>
  mutate(income_level = factor(
    income_level,
    levels = c("Low income", "Lower middle income",
               "Upper middle income", "High income"),
    ordered = TRUE
  ))

income_long |>
  summarise(
    n           = n(),
    n_missing   = sum(is.na(income_level)),
    pct_missing = round(100 * mean(is.na(income_level)), 1),
    first_year  = min(year[!is.na(income_level)]),
    last_year   = max(year[!is.na(income_level)])
  ) |>
  knitr::kable()
```

There is no missing data from 1987 to 2022, the years the World Bank classifies. Earlier years will appear missing once this is joined with longer variables, which is structural.

```{r}
income_latest <- income_long |>
  filter(!is.na(income_level), year == max(year[!is.na(income_level)]))

income_latest |>
  count(income_level) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  knitr::kable(caption = "World Bank income group, 2022")
```

```{r}
#| fig-cap: "Number of countries by income group, 2022"
ggplot(income_latest, aes(income_level)) +
  geom_bar(fill = "orange") +
  scale_x_discrete(labels = scales::label_wrap(12)) +
  labs(x = "Income group", y = "Number of countries") +
  theme_minimal()
```

```{r}
#| fig-cap: "Share of countries in each income group, 1987 to 2022"
income_long |>
  filter(!is.na(income_level)) |>
  count(year, income_level) |>
  group_by(year) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(year, pct, fill = income_level)) +
  geom_area() +
  labs(x = "Year", y = "Share of countries", fill = "Income group") +
  theme_minimal()
```

In 2022 High income is the largest group with 60 countries, or 30.9%, followed by Lower middle at 55, Upper middle at 53 and Low income at 26. Since 1987 the High income share rose from 18% to almost 31%, while Low income fell from 26% to 13%.

```{r}
#| fig-cap: "Population by income group, 2022"
income_pop_2022 <- income_latest |>
  inner_join(population_long |> filter(year == 2022),
             by = c("geo", "name", "year"))

income_pop_2022 |>
  group_by(income_level) |>
  summarise(n = n(), median_pop = median(population), mean_pop = mean(population)) |>
  knitr::kable(digits = 0, caption = "Population by income group, 2022")

ggplot(income_pop_2022, aes(income_level, population)) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.6, color = "orange", size = 2) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5,
               color = "black", linewidth = 0.4) +
  scale_x_discrete(labels = scales::label_wrap(12)) +
  scale_y_log10(labels = scales::comma) +
  labs(x = "Income group", y = "Population, log scale") +
  theme_minimal()
```

Median population is highest in Low income at 21.5 million and lowest in High income at 5.3 million, and every group mean exceeds its median. In Lower middle income the mean of 59 million is over four times the median of 13.8 million, driven by India and Pakistan. China and Indonesia do the same in Upper middle income. Population size and income group look largely unrelated.
